---
title: "Air quality compliance"
format: 
  html:
    self-contained: true
    toc: true
    toc-title: Contents
    toc-location: right
    toc-depth: 4
    theme: materia
    code-overflow: wrap
knitr: 
  opts_chunk:
    warning: false
    message: false   
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse) ; library(reactable)

# Nitrogen Dioxide annual mean Zone Agglomeration 2022
# Source: DEFRA
# URL: https://compliance-data.defra.gov.uk/datasets/Defra::nitrogen-dioxide-annual-mean-zone-agglomeration-2022/explore
no2_annual <- read_csv("data/Nitrogen_Dioxide_annual_mean_Zone_Agglomeration_2022.csv")

# Nitrogen Dioxide hourly mean Zone Agglomeration 2022
# Source: DEFRA
# URL: https://compliance-data.defra.gov.uk/maps/Defra::nitrogen-dioxide-hourly-mean-zone-agglomeration-2022/explore
no2_hourly <- read_csv("data/Nitrogen_Dioxide_hourly_mean_Zone_Agglomeration_2022.csv")

# PM10 annual mean Zone Agglomeration 2022
# Source: DEFRA
# URL: https://compliance-data.defra.gov.uk/maps/Defra::pm10-annual-mean-zone-agglomeration-2022/explore
pm10_annual <- read_csv("data/PM10_annual_mean_Zone_Agglomeration_2022.csv")

# PM10 daily mean Zone Agglomeration 2022
# Source: DEFRA
# URL: https://compliance-data.defra.gov.uk/datasets/pm10-daily-mean-zone-agglomeration-2022
pm10_daily <- read_csv("data/PM10_daily_mean_Zone_Agglomeration_2022.csv")

# PM2.5 annual mean Zone Agglomeration 2022
# Source: DEFRA
# URL: https://compliance-data.defra.gov.uk/datasets/pm2-5-annual-mean-zone-agglomeration-2022
pm25_annual <- read_csv("data/PM2_5_annual_mean_Zone_Agglomeration_2022.csv")

df <- bind_rows(no2_annual, no2_hourly, pm10_annual, pm10_daily, pm25_annual) %>% 
  mutate(pollutant = case_when(str_detect(pollutant_metric, "Nitrogen") ~ "NO2",
                               str_detect(pollutant_metric, "PM10") ~ "PM10",
                               str_detect(pollutant_metric, "PM2.5") ~ "PM2.5"),
         averaging_period = case_when(str_detect(pollutant_metric, "annual") ~ "Annual mean",
                                      str_detect(pollutant_metric, "hourly") ~ "1-hour mean",
                                      str_detect(pollutant_metric, "daily") ~ "24-hour mean"),
         uk_limit = case_when(pollutant == "NO2" & averaging_period == "Annual mean" ~ round(maximum_value/40,1),
                              pollutant == "PM10" & averaging_period == "Annual mean" ~ round(maximum_value/40,1),
                              pollutant == "PM2.5" & averaging_period == "Annual mean" ~ round(maximum_value/20,1)),
         who_limit = case_when(pollutant == "NO2" & averaging_period == "Annual mean" ~ round(maximum_value/10,1),
                               pollutant == "PM10" & averaging_period == "Annual mean" ~ round(maximum_value/15,1),
                               pollutant == "PM2.5" & averaging_period == "Annual mean" ~ round(maximum_value/5,1))) %>% 
  select(area_code = unique_code, area_name = name, pollutant, averaging_period, maximum_value, unit, uk_limit, who_limit)
```

The UK is divided into 43 zones for air quality assessment. These tables show how they fared under the [UK's annual air quality compliance assessment](https://uk-air.defra.gov.uk/air-pollution/uk-eu-limits) in 2022.

### Nitrogen Dioxide (N0<sub>2</sub>)

Zones must have an annual average nitrogen dioxide (N0<sub>2</sub>) level of 40 µg/m<sup>-3</sup> or below.
```{r}
filter(df, pollutant == "NO2" & averaging_period == "Annual mean") %>% 
  arrange(desc(uk_limit)) %>% 
  select(-c(area_code, pollutant, averaging_period)) %>% 
  reactable(columns = list(
    area_name = colDef(name = "Zone name"),
    maximum_value = colDef(name = "NO2"),
    unit = colDef(name = "Unit"),
    uk_limit = colDef(name = "No. times over legal limit"),
    who_limit = colDef(name = "No. times over WHO recommended levels")))
```

### Particulate Matter (PM<sub>10</sub>)

Zones must have an annual average particulate matter (PM<sub>10</sub>) level of 40 µg/m<sup>-3</sup> or below.
```{r}
filter(df, pollutant == "PM10" & averaging_period == "Annual mean") %>% 
  arrange(desc(uk_limit)) %>% 
  select(-c(area_code, pollutant, averaging_period)) %>% 
  reactable(columns = list(
    area_name = colDef(name = "Zone name"),
    maximum_value = colDef(name = "PM10"),
    unit = colDef(name = "Unit"),
    uk_limit = colDef(name = "No. times over legal limit"),
    who_limit = colDef(name = "No. times over WHO recommended levels")))
```

### Particulate Matter (PM<sub>2.5</sub>)

Zones must have an annual average particulate matter (PM<sub>2.5</sub>) level of 20 µg/m<sup>-3</sup> or below.
```{r}
filter(df, pollutant == "PM2.5" & averaging_period == "Annual mean") %>% 
  arrange(desc(uk_limit)) %>% 
  select(-c(area_code, pollutant, averaging_period)) %>% 
  reactable(columns = list(
    area_name = colDef(name = "Zone name"),
    maximum_value = colDef(name = "PM25"),
    unit = colDef(name = "Unit"),
    uk_limit = colDef(name = "No. times over legal limit"),
    who_limit = colDef(name = "No. times over WHO recommended levels")))
```

</br>
**Source**: [Department for Environment Food & Rural Affairs](https://compliance-data.defra.gov.uk/)